---
title: "Bootstrap Analysis with trieff"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bootstrap Analysis with trieff}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 4
)
```

```{r setup}
library(trieff)
library(dplyr)

# Initialize data once at the start
data <- init(sem_malmo_synth) %>% 
  sim_te()
```

## Introduction

Bootstrap analysis in trieff provides confidence intervals for Triage Effectiveness metrics and enables assessment of statistical stability. This vignette demonstrates how to use bootstrapping effectively and interpret the results.

The confidence intervals calculated here can be visualized using the plotting functions described in `vignette("plot_te")`. The visualization provides an intuitive way to understand the uncertainty in TE estimates.

Two bootstrapping methods are available in trieff:

1. **Standard Bootstrap**: Traditional approach that randomly samples individual cases with replacement.
2. **Segment Bootstrap**: Advanced approach that preserves temporal patterns by sampling segments of continuous queue activity.

Both methods work by repeatedly sampling from the dataset, calculating TE metrics for each sample, and using the distribution of these calculations to estimate confidence intervals.

## Prerequisites
- Knowledge on basic TE calculation (see `vignette("calc_te")`)
- Simulated theoretical wait times (see `vignette("sim_te")`)

## Why Bootstrap?

Bootstrapping helps us understand:

- The reliability of our TE metrics
- The stability of our estimates
- The impact of rare events and outliers

For reliable bootstrap analysis, we recommend using at least 3 months of ED data, though this guideline is based on preliminary analysis and may need adjustment for your specific context.

## Basic Bootstrap Usage

Let's start with a basic example using default bootstrap parameters. We'll use overall_only=TRUE to keep the output focused:

```{r basic-bootstrap}
# Calculate TE with bootstrap
te_results <- calc_te(data, 
                     bootstrap = TRUE, 
                     overall_only = TRUE)

# View results
print(te_results)
```

The output now includes confidence intervals for each TE metric, showing the uncertainty in our estimates.

## Advanced Bootstrap Configuration

The bootstrap process can be customized using several parameters:

```{r custom-bootstrap}
# Custom bootstrap settings
te_custom <- calc_te(data,
                     bootstrap = TRUE,
                     overall_only = TRUE,
                     seed = 123, # use seed to make your results reproducible
                     bootstrap_params = list(
                       sample_percentage = 0.8,  # Use 80% of data per iteration
                       n_iterations = 1000,      # decrease iterations
                       distribution_span = 0.99  # Wider confidence intervals
                     ))

print(te_custom)
```

Key parameters:

- sample_percentage: Controls resample size (default = 1)
- n_iterations: Number of bootstrap iterations (default = 2000)
- distribution_span: Width of confidence intervals (default = 0.95)

## Standard vs. Segment Bootstrap

The trieff package offers two different bootstrapping methods:

### Standard Bootstrap

The standard bootstrap method (`bootstrap = "standard"` or simply `bootstrap = TRUE`) randomly samples individual patient cases with replacement. This is the traditional bootstrap approach.

```{r standard-bootstrap}
# Standard bootstrap
standard_bootstrap <- calc_te(data,
                             bootstrap = "standard")

plot_te(standard_bootstrap)
```

### Segment Bootstrap

The segment bootstrap method (`bootstrap = "segment"`) preserves temporal patterns by sampling segments of continuous queue activity. This approach maintains the dependency structure between patients who were in the ED during the same time period.

```{r segment-bootstrap}
# Segment bootstrap
segment_bootstrap <- calc_te(data,
                            bootstrap = "segment")

plot_te(segment_bootstrap)
```


## Comparing Bootstrap Methods

While our theoretical understanding suggests segment bootstrapping should be preferred for emergency department data, our empirical testing has shown that both methods typically produce equivalent results in practice with comparable confidence intervals amd neither consistently producing wider or narrower intervals than the other. Despite theoretical differences in how they handle data dependencies, their practical outputs in terms of running means and confidence interval convergence are remarkably similar.
Both methods are maintained in the package to support methodological research and to allow users to validate results across different bootstrapping approaches, but for most practical applications, the standard bootstrap provides an efficient and reliable option.

```{r compare-methods}
# Function to extract confidence intervals
get_ci_width <- function(result) {
  ci_width <- result$results %>%
    mutate(ote_ci_width = boot_ote_var_upper - boot_ote_var_lower) %>%
    select(unit, ote_ci_width)
  return(ci_width)
}

# Compare CI widths
standard_ci <- get_ci_width(standard_bootstrap)
segment_ci <- get_ci_width(segment_bootstrap)

# Display results
comparison <- standard_ci %>%
  left_join(segment_ci, by = "unit", suffix = c("_standard", "_segment")) %>%
  mutate(difference = ote_ci_width_segment - ote_ci_width_standard)

print(comparison)
```


## How Segment Bootstrap Works

The segment bootstrap can be described by a custom block bootstrap that operates by:

1. Identifying segments of continuous queue activity (periods where the queue starts at 0 and ends at 0)
2. Sampling these segments with replacement
3. Calculating TE metrics for each resampled dataset

This approach preserves the temporal structure of patient flow, including:

- Arrival patterns

- Concurrent cases

- Crowding effects

- Priority interrelationships

While block bootstrapping reduces the effect of temporal effects, we argue that this approach should remove this effect altogether.


## Convergence Analysis

Convergence analysis helps verify that our bootstrap estimates are stable. When check_convergence=TRUE (which it is by standard), calc_te generates three types of diagnostic plots:

1. Running Mean Plot
   - Shows how the average TE estimate evolves with more iterations
   - Should stabilize to a roughly horizontal line
   - Large fluctuations at the end suggest need for more iterations

2. CI Width Plot
   - Shows how confidence interval width changes with iterations
   - Should decrease and stabilize
   - Continued narrowing suggests potential for more precise estimates

3. Standard Error Plot
   - Shows how the standard error of estimates changes
   - Should decrease and stabilize
   - Helps assess estimate precision

Let's examine convergence for our analysis:

```{r convergence, fig.width=10, fig.height=5, fig.alt="Convergence plots for the bootstrap."}
# Calculate TE with convergence checking
te_convergence <- calc_te(data,
                         bootstrap = TRUE,
                         overall_only = TRUE,
                         check_convergence = TRUE)

# View convergence plots
print(te_convergence$convergence)
```

Interpreting convergence plots:
- Look for stabilization in all three metrics
- Check if late iterations show systematic changes
- Verify CI width has reached reasonable stability

If convergence looks poor, consider:
- Increasing n_iterations
- Checking for data quality issues
- Verifying sufficient sample size

## Summary

Bootstrap analysis in trieff provides valuable insights into the reliability of TE metrics. Key points:

1. Use bootstrapping to get confidence intervals for TE metrics
2. Choose the appropriate bootstrap method:
   - Standard bootstrap for independent cases
   - Segment bootstrap for preserving temporal patterns
3. Verify convergence using diagnostic plots
4. Adjust parameters based on dataset size and precision needs
5. Consider computational resources for large datasets

Remember that while bootstrapping provides valuable statistical information, it should be combined with domain knowledge and practical considerations when evaluating triage system performance.
