---
title: "Calculating Triage Effectiveness"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculating Triage Effectiveness}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(trieff)
library(dplyr)
```

## Introduction

This vignette demonstrates how to calculate Triage Effectiveness (TE) using the `calc_te` function from the trieff package. We'll use the included `sem_malmo_synth` dataset, which contains synthetic emergency department visit data based on Malmö Hospital statistics.

It is recommended to run the simulation to generate Theoretical TE values before calculating TE, especially if you're interested in analyzing theoretical wait times (see the Simulation vignette for details).

## Prerequisites
- Initialized data (see `vignette("getting_started")`)

- Simulated theoretical wait times (see `vignette("sim_te")`) - not a strict prerequisite.


## Basic Usage

Let's start with a basic example of calculating TE. Note that by default, results are calculated for each medical unit separately since they typically operate as independent queues:

```{r basic-example}
# First initialize the data
data <- init(sem_malmo_synth)

# Run simulation to get theoretical wait times
data <- sim_te(data)

# Calculate TE metrics
te_results <- calc_te(data)

# Print results using the custom print method
print(te_results)
```

## Overall Results Only

When working with just the basic unit division, getting both unit-level and overall results is manageable. However, as we'll see later with subgroup analysis, the output can become quite extensive. The `overall_only` parameter can help manage this - let's introduce it here:

```{r overall-example}
# Calculate overall TE metrics
te_overall <- calc_te(data, overall_only = TRUE)

# Print results
print(te_overall)
```

## Subgroup Analysis

We can analyze specific subgroups, such as age groups. Notice how the output becomes quite extensive with results for each combination of unit and age group:

```{r subgroup-example}
# Define age groups
data <- data %>%
  mutate(age_group = cut(age_at_arrival, 
                        breaks = c(0, 59, 79, Inf),
                        labels = c("18-59", "60-79", "80+")))

# Analyze by age group
te_age <- calc_te(data, var1 = "age_group")

# Print results
print(te_age)
```

For a clearer overview when working with subgroups, we can use `overall_only` to focus on the main patterns:

```{r subgroup-overall}
# Analyze by age group, overall results only
te_age_overall <- calc_te(data, var1 = "age_group", overall_only = TRUE)

# Print results
print(te_age_overall)
```

## Summary

This vignette has demonstrated the main features of the `calc_te` function:

1. Basic TE calculation by unit
2. Overall TE calculation
3. Subgroup analysis with both detailed and overall results

For more advanced topics such as bootstrap analysis and convergence checking, please see the Bootstrap and Convergence vignette. For visualizing these results, see the Plotting vignette.

For more details on specific parameters and advanced usage, refer to the function documentation (`?calc_te`) and the instructions for bootstrapping (`?bootstrap`).
