---
title: "Heatmap Analysis in trieff"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Heatmap Analysis in trieff}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 8
)
```

```{r setup}
library(trieff)
library(dplyr)
```

## Introduction

The heatmap functionality in trieff is primarily a research tool designed to explore and validate the theoretical properties of Triage Effectiveness (TE). Unlike other features in the package, heatmap generation is computationally intensive and typically used for understanding fundamental relationships rather than day-to-day analysis.

## Prerequisites
- Understanding of TE concepts (`vignette("getting_started")`)

- Familiarity with simulation (`vignette("sim_te")`)

- Could require some computational resources.


## Relationship to Simulation
The heatmap analysis builds on the simulation framework described in `vignette("sim_te")`. However, it requires significantly more computational resources due to repeated simulations across sensitivity/specificity combinations.

### Important Considerations

1. **Computational Cost**: Full heatmap generation with typical ED datasets (>100,000 visits) can take several minutes depending on step size.
2. **Intended Use**: This tool is meant for researchers validating TE properties or comparing results across different ED settings.
3. **Limitation**: The heatmap does not naturally handle different units and it is recommended to only simulate one unit per heatmap.

## Understanding Heatmap Results

A TE heatmap visualizes how different combinations of sensitivity and specificity affect TE values. The key patterns we have established so far are:

1. The sum of sensitivity and specificity must exceed 1 for positive TE values
2. When this sum exceeds 1, improving sensitivity yields slightly better gains
3. When below 1, improving specificity is more beneficial

Let's examine these patterns with a small dataset:

```{r example of heatmap, fig.width=7, fig.height=7}
# Get a small sample (1 week) of data
small_data <- init(sem_malmo_synth[
  sem_malmo_synth$arrival >= min(sem_malmo_synth$arrival) &
  sem_malmo_synth$arrival <= min(sem_malmo_synth$arrival) + lubridate::days(7),
])

data <- sem_malmo_synth %>%
    filter(unit == "medical") %>% 
    init()
# As noted before - use only one unit per sim.

# Generate heatmap data
# WARNING: Even with this small dataset, this may take several minutes
heatmap_data <- data %>% 
  sim_heat(n_workers = 12, step_size = 10)

plot_te_heatmap(heatmap_data) # Create heatmap
```

We can also normalize the heatmap with a single argument:

```{r example of heatmap - normalized, fig.width=7, fig.height=7}
plot_te_heatmap(heatmap_data, normalize = TRUE) # Create heatmap

```
## Alternative Calculations

The package includes several alternative TE calculations (cube root, log, median-based) primarily for research validation. These can be accessed using `alt_calc = TRUE` in `sim_heat_alt()`, i.e. it is not possible based on the `sim_heat()`:

```{r example alt_calc, fig.width=10, fig.height=5}
# Example of alternative calculations
sim_heat_alt(data, alt_calc = TRUE, step_size = 25) %>% 
  plot_te_heatmap(show_alt_calc = TRUE)
```

These alternatives help validate that the patterns we observe aren't artifacts of our calculation method.

## Summary

Heatmap analysis in trieff is a powerful research tool for understanding TE properties, but it requires:
1. Significant computational resources/patience
2. Careful interpretation of results
3. Understanding of the theoretical background

For regular TE analysis of your ED, we recommend using the main `calc_te()` function as demonstrated in other vignettes.

For researchers interested in detailed heatmap analysis, we recommend reviewing the methods section of the TE paper and ensuring sufficient computational resources before beginning analysis.
