---
title: "Heatmap Analysis in trieff"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Heatmap Analysis in trieff}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 8
)
```

```{r setup}
library(trieff)
library(dplyr)
library(patchwork)
```

## Introduction

The heatmap functionality in trieff is primarily a research tool designed to explore and validate the theoretical properties of Triage Effectiveness (TE). Unlike other features in the package, heatmap generation is computationally intensive and typically used for understanding fundamental relationships rather than day-to-day analysis.

trieff supports both Waiting Time-based TE (WTE) and Rank-based TE (RTE) methods for heatmap generation, each offering different perspectives on triage system performance. However RTE is naturaly normalizes between 0% and 100% which makes its results more directly comparable across different ED settings and sensitivity/specificity combinations.

## Prerequisites
- Understanding of TE concepts (`vignette("getting_started")`)
- Familiarity with simulation (`vignette("sim_te")`)
- Understanding of WTE and RTE differences (`vignette("calc_wte_rte")`)
- Computational resources for processing


## Relationship to Simulation
The heatmap analysis builds on the simulation framework described in `vignette("sim_te")`. However, it requires significantly more computational resources due to repeated simulations across sensitivity/specificity combinations.

### Important Considerations
1. **Computational Cost**: Full heatmap generation with typical ED datasets (>100,000 visits) can take several minutes up to hours depending on step size.
2. **Intended Use**: This tool is meant for researchers validating TE properties or comparing results across different ED settings.
3. **Limitation**: The heatmap does not naturally handle different units and it is recommended to only simulate one unit per heatmap.

## Understanding Heatmap Results

A TE heatmap visualizes how different combinations of sensitivity and specificity affect TE values. The key patterns we have established so far are:

1. The sum of sensitivity and specificity must exceed 1 for positive TE values
2. When this sum exceeds 1, improving sensitivity yields slightly better gains
3. When below 1, improving specificity is more beneficial
4. Above rules are generally hard to validate when the sum is very close to 1, likely due to stochastic reasons
5. If inflow of LOSET patients are uneven during operational hours and the mean waiting time is also uneven, such as that more time critical cases arrives during the night when waiting times are longer the values where the sensitivity specificity sum is close to 1 can diverge further from 0 for WTE, but not RTE.

Let's compare WTE and RTE heatmaps with a small dataset:

```{r example_of_heatmaps, fig.width=12, fig.height=6, fig.alt="Comparison of WTE and RTE heatmaps showing TE over sensitivity/specificity combinations"}

data <- load_sem_synth() %>%
    filter(unit == "C") %>% # Use only one unit per sim
    filter(arrival < min(arrival) + lubridate::days(30)) %>% # reducing data size for speed, don't do this for real analysis :)
    init()

# Generate heatmap data using both calculation methods
# WARNING: Even with this small dataset, this may take several minutes
wte_heatmap <- data %>% 
  sim_heat(n_workers = 4, step_size = 10, calc_method = "wte")

rte_heatmap <- data %>% 
  sim_heat(n_workers = 4, step_size = 10, calc_method = "rte")

# Create and combine heatmaps
wte_plot <- plot_te_heatmap(wte_heatmap, title = "WTE Heatmap")
rte_plot <- plot_te_heatmap(rte_heatmap, title = "RTE Heatmap")

# Display side by side
wte_plot | rte_plot
```

Notice how both heatmaps reveal similar patterns, but the RTE heatmap (right) may show a more consistent scale due to its natural normalization properties.

## Normalization in Heatmaps

Normalization is particularly important when comparing heatmaps across different EDs or datasets. While RTE is naturally normalized due to its definition, WTE may benefit from explicit normalization to facilitate comparison:

```{r example_of_heatmap_normalized, fig.width=12, fig.height=6, fig.alt="Comparison of WTE and RTE heatmaps with WTE normalized"}
# Create normalized WTE heatmap
wte_norm_plot <- plot_te_heatmap(wte_heatmap, normalize = TRUE, 
                             title = "WTE Heatmap (Normalized)")

# RTE doesn't typically require normalization
rte_norm_plot <- plot_te_heatmap(rte_heatmap, 
                             title = "RTE Heatmap (No Normalization Needed)")

# Display side by side
wte_norm_plot | rte_norm_plot
```

Normalization in WTE scales values to make them more visually comparable. For RTE, normalization is less necessary since the values are already on a scale that naturally runs from negative values (worse than random) through 0 (equivalent to random) to 1 (perfect prioritization).

## Color Scheme Options

The `plot_te_heatmap` function supports different color schemes through the `color_scheme` parameter. This is particularly useful for choosing appropriate visualizations for different audiences, including options that are colorblind-friendly.

### Available Color Schemes

```{r color_schemes, fig.width=12, fig.height=10, fig.alt="Four heatmaps showing the same TE data with different color schemes"}
# Create a 2x2 grid of heatmaps with different color schemes
color_schemes <- c("oceansky", "blueorange", "thermalsafe", "triage")

# Generate plots with each color scheme
color_scheme_plots <- list()

for (scheme in color_schemes) {
  color_scheme_plots[[scheme]] <- plot_te_heatmap(
    rte_heatmap, # Using RTE heatmap for color scheme examples
    color_scheme = scheme,
    title = paste(toupper(substr(scheme, 1, 1)), substr(scheme, 2, nchar(scheme)), "Scheme", sep="")
  )
}

# Combine the plots in a grid
patchwork::wrap_plots(color_scheme_plots, ncol = 2)
```

Each color scheme has different characteristics:

**OceanSky** (default): Uses medium blue (#3a75c4), light blue, light yellow, and gold. Fully colorblind-safe with excellent text readability across all colors. Provides an intuitive "cool-to-warm" progression that works well for diverse audiences.

**BlueOrange**: Implements a scientifically-designed diverging color palette based on ColorBrewer's "RdBu" (Red-Blue) scale (#2c7bb6, #abd9e9, #fdae61, #d7191c). 

**ThermalSafe**: Uses purple, light blue, yellow, and orange. Colorblind-safe and evokes thermal imaging aesthetics. Provides strong differentiation between values while maintaining accessibility.

**Triage**: Uses classic triage colors (red, orange, yellow, green) familiar in healthcare settings. Not colorblind-safe but immediately recognizable to healthcare professionals who work with triage systems.


## Summary

Heatmap analysis in trieff is a powerful research tool for understanding TE properties. Key takeaways:

1. Both WTE and RTE calculation methods are available for heatmap generation
2. RTE offers naturally normalized results, making it well-suited for comparative analysis
3. WTE may require explicit normalization for optimal visualization
4. Both methods reveal similar fundamental patterns in the relationship between sensitivity, specificity, and TE
5. Use the `calc_method` parameter in `sim_heat()` to select your preferred calculation approach

For researchers interested in detailed heatmap analysis, we recommend reviewing the methods section of the TE paper (see `vignette("original_study")`) and ensuring sufficient computational resources before beginning analysis.

For regular TE analysis of your ED, we recommend using the main `calc_wte()` or `calc_rte()` functions as demonstrated in other vignettes.
