---
title: "Plotting Triage Effectiveness Results"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plotting Triage Effectiveness Results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 4
)
```

```{r setup}
library(trieff)
library(dplyr)

# Initialize and simulate data once at the start
data <- init(sem_malmo_synth) %>% 
  sim_te()
```

## Introduction

The `plot_te()` function provides flexible visualization of Triage Effectiveness results. This vignette demonstrates how to create and customize these visualizations, building on the analyses shown in the calc_te and bootstrap vignettes.

## Prerequisites
- Calculated TE metrics (see `vignette("calc_te")`)
- Optional: Bootstrap results for confidence intervals

## Basic Plotting

Let's start with a basic overall TE plot:

```{r basic-plot, fig.width=7, fig.height=1.2, fig.alt="A simple Triage Effectiveness plot"}
# Calculate TE metrics
te_results <- calc_te(data, overall_only = TRUE)

# Create basic plot
plot_te(te_results)
```

By default, the plot shows both Theoretical TE (teal) and Observed TE (purple) values. The horizontal lines connecting these points represent the Observed-Theoretical Gap (OTG).

## Plotting Bootstrap Results

The `plot_te()` function automatically handles bootstrap results when they're available. Let's see how this changes the visualization:

```{r bootstrap-plot, fig.width=7, fig.height=1.2, fig.alt="Triage Effectiveness plot with bootstrap"}
# Calculate TE with bootstrap
te_bootstrap <- calc_te(data,
                       bootstrap = TRUE,
                       overall_only = TRUE)

# Plot with confidence intervals
plot_te(te_bootstrap)
```

Notice how the plot now includes variability intervals (lighter shaded areas) for both TTE and OTE. These intervals are derived from the bootstrap analysis and help visualize the uncertainty in our estimates.

## Subgroup Analysis

We can visualize TE patterns across subgroups. Let's use the age group example from the calc_te vignette:

```{r subgroup-plot, fig.width=7, fig.height=2, fig.alt="Triage Effectiveness based on subgroups with bootstrapping"}
# Create age groups
data <- data %>%
  mutate(age_group = cut(age_at_arrival, 
                        breaks = c(0, 59, 79, Inf),
                        labels = c("18-59", "60-79", "80+")))

# Calculate and plot TE by age group
te_age <- calc_te(data,
                  var1 = "age_group",
                  bootstrap = TRUE,
                  overall_only = TRUE)

plot_te(te_age)
```

## Handling Negative TE Values

In some cases, particularly when analyzing specific units, we might encounter negative TE values. This requires special handling of the plot's x-axis. Let's demonstrate this with two plots - one showing only orthopedic units (which in the SEM synthpop has negative values) and another showing all units:

```{r negative-te-1, fig.width=7, fig.height=1.2, fig.alt="A Triage Effectiveness plot that has negative values naturally"}
# First, let's get results for orthopedic unit only
ortho_data <- data %>%
  filter(unit == "orthopedics")

te_ortho <- calc_te(ortho_data, bootstrap = TRUE)

data_other <- data %>% 
    filter(!unit == "orthopedics")  

# Then get results for all units minus orthopedics
te_all <- calc_te(data_other, bootstrap = TRUE)

# Plot orthopedic results
plot_te(te_ortho)
```

```{r negative-te-2, fig.width=7, fig.height=2, fig.alt="A Triage Effectiveness plot that has negative values forced"}

# Plot all units with adjusted x-axis
plot_te(te_all, min_x = -0.4)
```

Notice how setting `min_x = -0.3` ensures consistent scale and alignment between the plots, making it easier to compare results across different units.

## Advanced Plot Customization

While the default settings are suitable for most purposes, `plot_te()` offers several customization options:

```{r customization, fig.width=7, fig.height=1.2, fig.alt="A Triage Effectiveness plot with custom arguments changing its appearance"}
# Example with custom settings
plot_te(te_bootstrap,
        var_alpha = 0.9,          # less transparent variability intervals
        dumbell_width = 1.5,      # Thicker connecting lines
        label_style = "small",    # small text labels
        title = "Custom TE Plot") # Custom title
```

Key customization parameters include:

- `var_alpha`: Controls transparency of variability intervals (0-1)

- `dumbell_width`: Width of lines connecting TTE and OTE points

- `label_style`: Can be "none", "small", "full", or "half" (default)

- `show_tte`/`show_ote`: Toggle display of specific metrics

- `show_var`: Toggle display of variability intervals

For more details on available parameters, see `?plot_te`.

## Summary

This vignette has demonstrated how to:

1. Create basic TE plots

2. Visualize bootstrap confidence intervals

3. Handle subgroup analyses

4. Deal with negative TE values

5. Customize plot appearance

For more complex analyses, consider combining these visualization techniques with the methods described in the calc_te and bootstrap vignettes.
