---
title: "Plotting Triage Effectiveness Results"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plotting Triage Effectiveness Results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 4
)
```

```{r setup}
library(trieff)
library(dplyr)
library(patchwork)

# Initialize and simulate data once at the start
data <- init(sem_malmo_synth) %>% 
  sim_te()
```

## Introduction

The `plot_te()` function provides flexible visualization of Triage Effectiveness results. This vignette demonstrates how to create and customize these visualizations for both Waiting Time-based TE (WTE) and Rank-based TE (RTE) calculations.

## Prerequisites
- Calculated TE metrics using `calc_wte()` or `calc_rte()` (see `vignette("calc_wte_rte")`)
- Optional: Bootstrap results for confidence intervals (see `vignette("bootstrap")`)

## Understanding TE Visualization Methods

Before diving into examples, it's important to understand the two complementary approaches to measuring and visualizing Triage Effectiveness:

1. **Waiting Time-based TE (WTE)**: Measures how well triage reduces the actual waiting time for time-critical patients.

2. **Rank-based TE (RTE)**: Evaluates how effectively triage improves queue positions for time-critical patients.

Both approaches have their strengths and different confidence interval calculation methods, which will be demonstrated throughout this vignette.

## Basic Plotting

Let's start with basic plots for both WTE and RTE:

```{r basic-plots, fig.width=10, fig.height=3, fig.alt="Comparison of WTE and RTE plot visualizations"}
# Calculate WTE metrics
wte_results <- calc_wte(data, overall_only = TRUE)

# Calculate RTE metrics 
rte_results <- calc_rte(data, overall_only = TRUE)

# Create basic plots
wte_plot <- plot_te(wte_results, title = "Waiting Time-based TE (WTE)")
rte_plot <- plot_te(rte_results, title = "Rank-based TE (RTE)", ci_vars_override = "none")

# Display plots
wte_plot / rte_plot
```

By default, the plots show both Theoretical TE (teal) and Observed TE (purple) values. The horizontal lines connecting these points represent the Observed-Theoretical Gap (OTG).

## Confidence Interval Types

The `plot_te()` function offers several options for displaying confidence intervals using the `ci_vars_override` parameter. This is particularly important when working with RTE, as queue positions often have non-normal distributions.

The available confidence interval types are:

1. **Parametric**: Based on normal distribution assumptions (t-distribution)
2. **Non-parametric**: Based on quantiles of the actual distribution
3. **Bootstrap**: Derived from bootstrap resampling
4. **None**: No confidence intervals displayed

Let's see how these different CI types affect our visualizations:

```{r ci-types, fig.width=10, fig.height=6}
# Calculate RTE with all CI types available
rte_boot <- calc_rte(data, bootstrap = TRUE, overall_only = TRUE)

# Create plots with different CI types
plot1 <- plot_te(rte_boot, ci_vars_override = "parametric", 
                title = "RTE with Parametric CI")

plot2 <- plot_te(rte_boot, ci_vars_override = "nonparametric", 
                title = "RTE with Non-parametric CI")

plot3 <- plot_te(rte_boot, ci_vars_override = "boot", 
                title = "RTE with Bootstrap CI")

plot4 <- plot_te(rte_boot, ci_vars_override = "none", 
                title = "RTE with No CI")

# Display plots
(plot1 / plot2) | (plot3 / plot4)
```

Note how the non-parametric confidence intervals are often wider than parametric ones for RTE due to the non-normal distribution of queue positions. For most analyses, bootstrap confidence intervals provide a good balance of validity and interpretability.

## Automatic CI Selection

The `plot_te()` function intelligently selects the most appropriate confidence interval type based on a priority order:

1. User-specified override via `ci_vars_override` parameter
2. Bootstrap results (if available)
3. Recommended CIs based on calculation method and normality tests:
   - For RTE: based on Shapiro-Wilk test results
   - For WTE: bootstrap CIs are recommended when available
4. None (if no other options are available)

This automatic selection ensures appropriate uncertainty representation without requiring manual specification in most cases.

## Bootstrap Confidence Intervals

Bootstrap confidence intervals are available for both WTE and RTE when calculated with the `bootstrap = TRUE` parameter:

```{r bootstrap-plots, fig.width=10, fig.height=3}
# Calculate TE with bootstrap for both methods
wte_bootstrap <- calc_wte(data, bootstrap = TRUE, overall_only = TRUE)
rte_bootstrap <- calc_rte(data, bootstrap = TRUE, overall_only = TRUE)

# Plot with bootstrap confidence intervals
wte_boot_plot <- plot_te(wte_bootstrap, title = "WTE with Bootstrap CI")
rte_boot_plot <- plot_te(rte_bootstrap, title = "RTE with Bootstrap CI")

# Display plots
wte_boot_plot / rte_boot_plot
```

## Subgroup Analysis

Both WTE and RTE support subgroup analysis. Let's create age groups and visualize TE patterns:

```{r subgroup-plots, fig.width=10, fig.height=6}
# Create age groups
data <- data %>%
  mutate(age_group = cut(age_at_arrival, 
                        breaks = c(0, 59, 79, Inf),
                        labels = c("18-59", "60-79", "80+")))

# Calculate and plot TE by age group
wte_age <- calc_wte(data, var1 = "age_group", bootstrap = TRUE, overall_only = TRUE)
rte_age <- calc_rte(data, var1 = "age_group", bootstrap = TRUE, overall_only = TRUE)

# Create plots
wte_age_plot <- plot_te(wte_age, title = "WTE by Age Group")
rte_age_plot <- plot_te(rte_age, title = "RTE by Age Group")

# Display plots
wte_age_plot / rte_age_plot
```

## Handling Negative TE Values

In some cases, particularly when analyzing specific units, we might encounter negative TE values. The `min_x` parameter allows standardization of the x-axis range across multiple plots for easier comparison:

```{r negative-te, fig.width=10, fig.height=6}
# Get data for different units
ortho_data <- data %>% 
  filter(unit == "orthopedics") 
medical_data <- data %>% 
  filter(unit == "medical") 

# Calculate WTE for both units
wte_ortho <- calc_wte(ortho_data, bootstrap = TRUE)
wte_medical <- calc_wte(medical_data, bootstrap = TRUE)

# Plot with default scaling - notice the different axes
plot1 <- plot_te(wte_ortho, title = "Orthopedics (default scaling)")
plot2 <- plot_te(wte_medical, title = "Medical (default scaling)")

# Plot with standardized x-axis
plot3 <- plot_te(wte_ortho, min_x = -0.4, title = "Orthopedics (standardized x-axis)")
plot4 <- plot_te(wte_medical, min_x = -0.4, title = "Medical (standardized x-axis)")

# Display plots showing the difference
(plot1 / plot2) | (plot3 / plot4)
```

Setting `min_x` ensures consistent scale and alignment between plots, making it easier to compare results across different units or groups.

## Advanced Plot Customization

The `plot_te()` function offers several customization options to tailor visualizations to specific needs. Let's explore how different parameters affect the appearance:

```{r customization-label-styles, fig.width=10, fig.height=8}
# Base plot for comparison
base_plot <- plot_te(wte_bootstrap, title = "Default label style ('half')")

# Different label styles
small_labels <- plot_te(wte_bootstrap, 
                      label_style = "small", 
                      title = "Small labels")

full_labels <- plot_te(wte_bootstrap, 
                     label_style = "full", 
                     title = "Full labels")

no_labels <- plot_te(wte_bootstrap, 
                   label_style = "none", 
                   title = "No labels")

# Display label style variations
base_plot / small_labels / full_labels / no_labels
```

```{r customization-variability, fig.width=10, fig.height=6}
# Variability display options
more_transparent <- plot_te(wte_bootstrap, 
                          var_alpha = 0.3, 
                          title = "More transparent variability (alpha = 0.3)")

less_transparent <- plot_te(wte_bootstrap, 
                          var_alpha = 0.8, 
                          title = "Less transparent variability (alpha = 0.8)")

# Display variability transparency variations
more_transparent / less_transparent
```

```{r customization-line-widths, fig.width=10, fig.height=6}
# Line width options
thin_lines <- plot_te(wte_bootstrap, 
                    dumbell_width = 0.5, 
                    var_width = 0.8, 
                    var_end_width = 0.5, 
                    title = "Thinner lines")

thick_lines <- plot_te(wte_bootstrap, 
                     dumbell_width = 2, 
                     var_width = 2.5, 
                     var_end_width = 1.5, 
                     title = "Thicker lines")

# Display line width variations
thin_lines / thick_lines
```

```{r customization-selective-display, fig.width=10, fig.height=6}
# Selective display of metrics
tte_only <- plot_te(wte_bootstrap, 
                  show_tte = TRUE, 
                  show_ote = FALSE, 
                  title = "TTE only")

ote_only <- plot_te(wte_bootstrap, 
                  show_tte = FALSE, 
                  show_ote = TRUE, 
                  title = "OTE only")

# Display selective metrics
tte_only / ote_only
```

Key customization parameters include:

- `var_alpha`: Controls transparency of variability intervals (0-1)
- `dumbell_width`: Width of lines connecting TTE and OTE points
- `var_width`: Width of variability interval main lines
- `var_end_width`: Width of variability interval end caps
- `var_end_height`: Height of variability interval end caps
- `label_style`: Can be "none", "small", "full", or "half" (default)
- `show_tte`/`show_ote`: Toggle display of specific metrics
- `show_var`: Toggle display of variability intervals
- `ci_vars_override`: Control which confidence interval type to display

For more details on available parameters, see `?plot_te`.

## Summary

This vignette has demonstrated how to:

1. Create basic TE plots for both WTE and RTE
2. Understand and control confidence interval types
3. Visualize bootstrap confidence intervals
4. Handle subgroup analyses
5. Deal with negative TE values
6. Customize plot appearance

For more complex analyses, consider combining these visualization techniques with the methods described in the `calc_wte_rte` and `bootstrap` vignettes.
