---
title: "Simulation in trieff"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation in trieff}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(trieff)
library(dplyr)
```

## Introduction

The trieff package uses simulation as a prestep to the Theoretical Triage Effectiveness (TTE) calculation by modeling how patients would be processed if strictly following their assigned priority levels. This vignette explains how the simulation works and demonstrates its implementation.

## Understanding the Simulation

### Key Concepts

The simulation calculates theoretical wait times that would occur if:

1. Patients were seen strictly in order of priority

2. Within the same priority level, patients were seen in order of arrival

3. No post-triage priority changes occurred

### Optimized Implementation

The default simulation (sim_te) optimizes performance by focusing only on segments containing time-critical (LOSET-positive) cases. This optimization works because:

- For WTE calculations, we only need to determine the waiting times of LOSET-positive patients relative to the overall mean waiting time. The overall mean can be directly calculated from observed wait times.

- For RTE calculations, we similarly only need to determine the queue positions of LOSET-positive patients. Non-LOSET segments have no impact on the RTE metrics.

This selective simulation approach dramatically reduces computation time without affecting the accuracy of TE results. For segments without LOSET cases, the function simply uses the observed wait times, which produces mathematically equivalent results to fully simulating those segments.
For research purposes where complete queue behavior analysis is needed, the package also provides sim_te_alt which simulates waiting times for all patients in all segments.

## Basic Usage

Let's start with a basic example:

```{r basic-example}
# Initialize data
data <- init(sem_malmo_synth)

# Run simulation
simulated_data <- sim_te(data)

# Check first few rows
simulated_data %>% 
  select(loset, observed_wait_time, theoretical_wait_time) %>% 
  head(10)

```

## Simulation Parameters

The simulation accepts several parameters:

```{r parameters}
simulated_data <- sim_te(
  data,
  tte = TRUE,     # Calculate theoretical wait times (default)
  btte = FALSE,   # Skip binary theoretical calculations
  n_workers = 1   # Number of cores for parallel processing, generally this is faster on 1 core.
)
```

### Full Simulation Option

For research purposes or when all theoretical wait times are needed, you can use `sim_te_alt`:

```{r full-simulation}
# Run full simulation
full_sim_data <- sim_te_alt(data)

# First let's show that the mean observed and theoretical wait time is the same
print(mean(full_sim_data$observed_wait_time))
print(mean(full_sim_data$theoretical_wait_time))

# Compare results
full_sim_data %>% 
  select(loset, observed_wait_time, theoretical_wait_time) %>% 
  head(10)

```

The full simulation calculates theoretical wait times for all cases but takes slightly longer to run. A sharp eye might notice that the wait times look different than what is shown in the example above. This is because in sim_te (not the alt version) the theoretical_waiting time is set to the observed_waiting time for any segment without LOSET cases that are not simulated. This keeps the mean waiting time identical in the both columns.

## Performance Considerations

The optimized simulation offers significant performance benefits:

1. Reduced computation by focusing on relevant segments

2. Lower memory usage

3. Faster parallel processing

However, both approaches produce identical TE values:

```{r compare-results}
# Calculate TE with both methods
te_opt <- calc_wte(simulated_data, overall_only = TRUE)
te_full <- calc_wte(full_sim_data, overall_only = TRUE)

# Compare results
print(te_opt)
print(te_full)
```

## When to Use Each Approach

Use `sim_te` (default) when:

- Calculating TE metrics

- Processing large datasets

Use `sim_te_alt` when:

- Researching queue behavior

- Analyzing non-LOSET wait patterns

- Validating simulation methods

## Technical Details

The simulation uses a non-preemptive approach where:

- Each physician contact becomes an available time slot

- Patients are queued by priority and arrival time

- Wait times are calculated from slot assignment

This deterministic approach ensures reproducible results while accurately modeling priority-based queuing.

## Summary

Understanding simulation in trieff helps you:

1. Choose the appropriate simulation method

2. Interpret results correctly

3. Optimize performance for your use case

For most applications, the default `sim_te` provides the most efficient path to accurate TE calculation.
